A state space model reveals the rich and varied dynamics of replay

### 2020-07-14 Article.tex
Update on Overleaf.

\documentclass[times, twoside]{zHenriquesLab-StyleBioRxiv}
\usepackage{blindtext}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{mathtools}
\usepackage{blkarray, bigstrut}
\usepackage{bbold}
\usepackage{amsmath}
\DeclareMathOperator*{\E}{\mathbb{E}}
\DeclareMathOperator*{\argmax}{arg\,max}
\newcommand{\norm}[1]{\left\lVert #1 \right\rVert}

@@

Next, we specify a dynamics model for how latent position---the "mental" position of the animal represented by the cells--- evolves over time (Figure \ref{1}C). We do this by defining a state transition matrix, which defines how the latent position can change from the previous time step. Previous findings suggest three distinct dynamics: stationary \cite{JaiDistincthippocampalcorticalmemory2017, FarooqEmergencepreconfiguredplastic2019}, smoothly moving \cite{DavidsonHippocampalReplayExtended2009}, and fragmented / jumpy \cite{PfeifferAutoassociativedynamicsgeneration2015}, which could correspond to both spatially incoherent representations and representations that jump from one position to another. We therefore developed state transition models to capture each of these dynamics.

@@

One of our criteria for a more optimal method is that it use all of the available spiking data. Using only clustered spikes discards all of the spike events that cannot be uniquely assigned to a putative single neuron, reducing very substantially the quality of the resulting decoding. We therefore adopted a "clusterless" approach which directly relates spikes and their multiunit spike waveform features to position without spike sorting, which we call clusterless decoding (see Methods). Clusterless decoding previously has been used to successfully identify theta sequences and replay sequences in the hippocampus (\cite{KloostermanBayesiandecodingusing2014, ChenTransductiveneuraldecoding2012,DengRapidclassificationhippocampal2016, KayConstantSubsecondCycling2020}). Applying a clusterless decoder to the same SWR, we get similar classification of the sequence (Figure \ref{2}B, D), both with 1D linearized position and 2D position (Figure 2-Supplemental 1B). The spatial extent of the event is longer and the estimate of the posterior probability of position is narrower for the clusterless model, however. This reflects the clusterless model's access to a larger pool of data that provides more information about the extent of the event, and more certainty in the latent position and the dynamic (Figure \ref{2}D vs C).

@@

Previous work focusing on spatially sequential replay reported that only a minority of events contain sequential spatial content [REFS]. We therefore asked what fraction of classified events contained spatially coherent content in any form, defined as a SWR containing any times with stationary, stationary-continuous mixture, or continuous dynamics (see methods). We found that 22478 of 23382 or 96\% of classified SWRs had some spatially coherent structure. We then asked what fraction of events contained spatially incoherent content, defined as a SWR containing any times with fragmented-continuous mixture or fragmented. We found that 3353 of 23382 or 14\% of classified SWRs had some spatially incoherent structure. This was consistent across animals (Figure \ref{3}F). We then asked what percentage of classified SWR contained continuous content, like would typically be analyzed when using the Bayesian decoder. Here our findings were consistent with previous reports: in our datasets 4511 of  23382 or 19\% of classified SWRs had some continuous content (Figure \ref{3}G). Thus, focusing on only continuous content misses a large fraction of events where there is evidence for spatially coherent but not continuous content.

@@

Our model is specified in the context of a marked point process associated with different dynamics. This specification allows us to not only decode events, but also to quantify the model's certainty in each position estimate at each moment in time given the model parameters. To do so we can compute the cumulative spatial extent of the 95\% highest posterior density (HPD) values of the posterior linear position. Larger values of the cumulative HPD indicate the model is less certain about position, because the HPD values are spread out over more of the track at a given time point, whereas lower values indicate that the model is more certain about the estimate of position, because the HPD values are more concentrated and cover less of the track. Thus, the HPD provides a complementary measures of spatial coherence, and evaluating it allows us to verify that the events we defined as spatially coherent correspond to events where there is high certainty around the position estimates.

We found that spatially coherent events did indeed have lower cumulative HPDs. Figure \ref{4}A and B show two example SWRs that were classified as having stationary and continuous dynamics, respectively. The bulk of the HPD values at each time step in these SWRs is concentrated in a small portion of the track and the cumulative HPD is low throughout the SWRs. In contrast, Figure \ref{4}C shows a SWR where the dynamics are fragmented and correspondingly, the HPD is much more spatially diffuse and the cumulative HPD is much higher. The HPD also provides insights into the moment-by-moment structure of each event, which can change over the time course of a SWR. An example of this change is shown in Figure \ref{4}D, where the cumulative HPD is small for most of the SWR until the end, where the uncertainty of position becomes much higher, reflecting a transition from a spatially coherent to a spatially incoherent representation. Overall, when we examined the average cumulative HPD for each SWR, grouped by dynamic, we found a clear bi-modal split between spatially congruent dynamics and spatially incongruent dynamics (Figure \ref{4}E). For the spatially congruent dynamics, the cumulative HPD was much lower than the spatially incongruent dynamics (median 24 cm, spatially congruent vs. median 238 cm, spatially incongruent).

@@

Our results indicate that most events with coherent spatial content are not continuous; that is, trajectories that move many times the running speed of the animal. We therefore examined these events in more details. We first noted that most of the SWRs (16454 of 23382 classified SWRs or 70\%) were well described by a single dynamic. Surprisingly, the most common SWRs were categorized exclusively as stationary-continuous mixtures (9860 of 23382 classified SWRs or 42\%, Figure \ref{5}A). These events contain representations that move at rates consistent with real-world movement (Fig. 1G), but were slightly shorter, on average, than events with continuous dynamics (median duration: stationary-continuous-mixture 72 ms vs. continuous 95 ms, Figure \ref{5}B). Nonetheless, both these slow-moving events and continuous events represented locations that were some distance away from the animal (mean trajectory distance from animal's position: stationary-continuous-mixture 51 cm vs. continuous 42 cm, Figure \ref{5}C). This indicates that the trajectories of these SWRs, like ones that have faster continuous dynamics, do not represent the animal's actual position.

The next most common category is those SWRs that were exclusively stationary (5149 of 23382 or 22\%). Unlike the stationary-continuous mixtures, most of these events represented a location close to the the animal's actual position (Figure \ref{5}C). There were, however, a number of stationary events  that represented positions far from the animal. We set a threshold of 30 cm distance from the animal to identify "non-local stationary trajectories" and found these trajectories in 1585 of 23382 (~7\%) of classified SWRs. About half of these (731 classified SWRs) were stationary throughout the duration of the SWR. These non-local stationary trajectories most commonly represented reward wells or choice points (Figure \ref{5}D), consistent with \cite{JaiDistincthippocampalcorticalmemory2017}.

Finally, we examined the overall activity levels across dynamics and found that all the dynamics had similar average multiunit spiking rates by tetrode (Figure \ref{5}E). This indicates that all dynamics, including the slower dynamics such as stationary and stationary-continuous mixtures and importantly not just the continuous dynamics, were driven by sustained spiking information.

\section*{Discussion}

Summary

Challenges the notion that replays consist only of high speed events.

adjudicates/clarifies results of Pfiefer and Foster and Stella et al. Brownian motion but couldn't tell the speeds. Pfiefer and Foster didn't know the percentage of events that acted this way

Why our results are different from the Bayesian decoder
seek to describe as many SWR events, clusterless decoding, more permissive model with different dynamics, explicit descriptive model instead of implicit noisy model.

relationship to Johnson and Redish 2008, Maboudi, model represents middle ground between HMM and typical Bayesian decoder. allows SWR to quickly be identified because of the discrete states. Can work in 2D. Can precisely identify timing of dynamics because of acausal decoding

Can extend to incorporate other previously experienced environments. Add prior on dynamics of speed

@@

\subsection*{Behavioral Task}
All animals performed a W-track spatial alternation task, which is described in detail in \cite{KarlssonAwakereplayremote2009}. In brief, each day, animals alternated between 20 minute rest epochs in a rest box and 15 minutes run epochs in the W-track. On the W-track, animals alternated between two types of runs: inbound and outbound. On inbound runs, the animal had to run from an outer arm to the center arm. On outbound runs, the animals had to run from the center arm to the least recently visited outer arm. Animals were rewarded for performing the task correctly. Only run epochs with at least 9 hippocampal principal cells that fired at least 100 spikes were included in the analysis.

\subsection*{Position of the Animal and Linearization}
The animal's 2D position was estimated from digital video (30 Hz) of two infrared diodes placed on the head stage preamplifiers using a semi-automated analysis. In order to decrease the time it takes to run the model, the 2D position of the animal was converted into a 1D position. This is done by first defining a 2D graph representation of the track, where edges correspond to segments of the W-track and nodes represent intersection points between those segments. Then, based on the algorithm in \cite{NewsonHiddenMarkovmap2009}, we use a Hidden Markov Model (HMM) to classify the most likely track segment the animal is on. Using the HMM takes into account the time dependent nature of the data and helps prevents sudden jumps from one track segment to another, which is particularly important near intersections. The observation model of the HMM is Gaussian and it models the likelihood of being on a track segment as the Gaussian distance to that segment with a 5 cm standard deviation. The state transition model is an empirically estimated state transition that changes with each time point that tries to ensure that the Euclidean distance between successive position estimates is similar to the shortest path distance along the graph between successive position estimates. A slight bias of 0.1 is given to the diagonal to encourage staying on the same track segment. The most likely track segment the animal is on is computed using the Viterbi algorithm. After finding the track segment that corresponds to each 2D position, the 2D position is projected onto the track segment. This allows us to define a distance from the center well in terms of shortest path length on the track, where 0 cm represents the center well position. The linear distance can then be converted into a linear position by assigning each track segment a position in 1D space. 15 cm gaps were placed between the center arm, left arm, and right arms in 1D space to prevent any smoothing done in the model from bleeding over inappropriately. The code used for linearization can be found at \url{https://github.com/Eden-Kramer-Lab/loren_frank_data_processing}.

\subsection*{Sorted Spikes, Multiunit Spikes and Waveform Features}
To obtain the neural data used for decoding, hippocampus electrical potentials were recorded at 30 kHz, referenced to a corpus callosum tetrode, and then digitally filtered between 600 Hz and 6 kHz. Events were considered threshold crossing events if any one of the four tetrode wires exceeded a 60 mV threshold. These multiunit threshold crossing events and the associated peak amplitudes of the tetrodes wires---defined as the electrical potential value at the time of the maximum amplitude at the threshold crossing---were used as data for the clusterless version of the model.

@@

Lastly, we evaluate the likelihood of the observations $p(O_{k} \mid x_{k}, I_{k})$ based on an encoding model fit during the encoding period. We assume the likelihood is the same for each dynamic $I_k$, so we only need to estimate $p(O_{k} \mid x_{k})$. It has been shown \cite{ZhangInterpretingNeuronalPopulation1998, BrownStatisticalParadigmNeural1998} that the likelihood with sorted spikes can be computed as:
$$
p(O_{k} \mid x_{k}) = p(\Delta N_{k}^{(1:C)} \mid x_{k}) \propto
\prod^{C}_{i=1} [\lambda_{i}(t_{k} \mid x_{k})\Delta_{k}]^{N_{t_{k}}^{i}} \\
    \exp[-\lambda_{i}(t_{k} \mid x_{k})\Delta_{k}]
$$
where $N_{t_{k}}^{i}$ represents a spike at time $t_k$ from cell $i$, $\Delta_k$ is the time bin size, and $\lambda_{i}(t_{k} \mid x_{k})$ is the instantaneous firing rate of cell $i$ given position $x_{k}$. $\lambda_{i}(t_{k} \mid x_{k})$ is the "place field" of the cell, which can be estimated via generalized linear model during the encoding period (which we explain further in the Model Implementation section).

Likewise, it has been shown \cite{ChenTransductiveneuraldecoding2012, DengRapidclassificationhippocampal2016, KloostermanBayesiandecodingusing2014} that the clusterless likelihood can be computed as: 
$$
p(O_{k} \mid x_{k}) = p(\Delta N_{k}^{(1:E)}, \vec{m_k} \mid x_{k}) \propto
\prod^{E}_{i=1} [\lambda_{i}(t_{k}, \vec{m}_k \mid x_{k})\Delta_{k}]^{N_{t_{k}}^{i}} \\
    \exp[-\Lambda_{i}(t_{k} \mid x_{k})\Delta_{k}]
$$
where $\lambda_{i}(t_{k}, \vec{m}_k \mid x_{k})$ is now a generalized firing rate that depends on an associated wave form features $\Vec{m}$ and $\Lambda_{i}(t_{k} \mid x_{k})$ is a marginal firing rate that is equivalent to a place field estimated on multiunit spikes. Both of these quantities can be estimated via kernel density estimation based on multiunit spikes and their associated waveform features observed in the encoding period.

\subsection*{Model Implementation}
In order to implement the model, we used a grid based approximation of the latent position $x_k$ that respected the geometry of the track. For 1D linearized positions, we discretized the position space based on the same track graph we used for linearization by finding bins less than or equal to the chosen position bin size for each edge of the graph. For 2D positions, we discretized the position space by binning 2D positions occupied by the animal with equal sized bins of the chosen position bin size, followed by morphological opening to get rid of any holes smaller than the bin size. This was done using the scipy image processing tools \cite{SciPy1.0ContributorsSciPyfundamentalalgorithms2020}. This grid based approximation allows us to use Reimann sums to approximate the integrals in the causal filter and acausal smoother equations. We also used this grid in combination with the track graph in order to construct the appropriate 1D random walk transition matrices that respected the track geometry. To do this, we inserted the position bin centers as nodes in the track graph, and then computed the shortest path distance between all pairs of position bin centers. We then evaluated a zero mean Gaussian with a variance of 6 cm on these distances to get the appropriate transition matrix.

\subsection*{Encoding - Sorted Spikes}
In order to encode how each cell's spiking activity relates to position (the place field), we fit a generalized linear model (GLM) with a Poisson response distribution to each cell's spiking activity during movement (defined as time periods when the running speed is greater than 4 cm/s). We estimate the parameters $\beta$, which consist of $\beta_{0}$, the average firing rate over time, and $\beta_{i}$, weights for third degree B-splines basis functions $f_{i}$ over position (or tensor products of the B-splines when position is two dimensional). B-spline basis functions are used because place field firing activity is assumed to vary smoothly over position and this prior knowledge can be exploited to reduce the total number of model parameters needed. Each basis function is spaced every 5 cm over the range of the position and zero constrained so that the change encoded by the parameters is relative to the baseline firing rate. We use a log link function to convert the linear combination of parameters to an instantaneous firing rate over time $\lambda(t)$ to ensure the rate is always positive.

@@

A small L2 penalization term $-\lambda\norm{\beta_{i}}_{2}^{2}$ used to prevent model fitting instability when spiking activity is very low. We set this to 0.5 for all cells. Fitting is done by maximizing the penalized likelihood using a Newton-Raphson algorithm. The code used to fit the GLMs is available at \url{https://github.com/Eden-Kramer-Lab/regularized_glm}.

\subsection*{Encoding - Clusterless}
In order to encode how each tetrode's unsorted spiking activity and waveform features relate to position, For each multiunit spike during movement (defined as time periods when the running speed is greater than 4 cm/s), the associated waveform features and position are recorded in a (n-spikes, n-features + n-position-dimensions) array. In our case, the waveform features correspond to the max amplitude observed on each tetrode wire at the time of the multiunit spike. This array is used as the training samples in a kernel density estimator during the decoding step.

Both of these rates can be defined as:

$$
\lambda_{i}(t_{k}, \vec{m}_k \mid x_{k}) = \mu_i \frac{p_{i}(x_k, \vec{m}_k)}{\pi(x)}
$$
and
$$
\Lambda_{i}(t_{k} \mid x_{k}) = \mu_i \frac{p_{i}(x_k)}{\pi(x)}
$$
where $\mu_i$ is the mean firing rate for tetrode $i$, $\pi(x)$ is the spatial occupancy of the animal on the track, and $p_{i}(x_k, \vec{m}_k)$ and $p_{i}(x_k)$ are the joint probability of observing waveform features $\Vec{m}$ and position $x$ and the probability of observing position $x$, respectively. $\pi(x)$, $p_{i}(x_k)$, and $p_{i}(x_k, \vec{m}_k)$ can all be estimated using the kernel density estimator:
$$
kde(y) = \frac{1}{N h_1...h_D} \sum^{N}_{i=1} \prod^{D}_{d=1} K_{d}\left(\frac{y_d - y_{i,d}}{h_d}\right)
$$
where $y$ is the data with $D$ dimensions, $K$ is a 1 dimensional Gaussian kernel with bandwidth $h_d$, $y_{i, d}$ is the data observed during the encoding period, $N$ is the number of observations in the encoding period.
\subsection*{Decoding}

position bin size 3 cm

\subsection*{Decoding - Sorted Spikes}

>>>>

\documentclass[times, twoside]{zHenriquesLab-StyleBioRxiv}
\usepackage{blindtext}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{mathtools}
\usepackage{blkarray, bigstrut}
\usepackage{bbold}
\usepackage{amsmath}
\DeclareMathOperator*{\argmax}{arg\,max}

@@

Next, we specify a dynamics model for how latent position---the "mental" position of the animal represented by the cells--- evolves over time (Figure \ref{1}C). We do this by defining a movement state transition matrix, which defines how the latent position can change from the previous time step. Previous findings suggest three distinct dynamics: stationary \cite{JaiDistincthippocampalcorticalmemory2017, FarooqEmergencepreconfiguredplastic2019}, smoothly moving \cite{DavidsonHippocampalReplayExtended2009}, and fragmented / jumpy \cite{PfeifferAutoassociativedynamicsgeneration2015}, which could correspond to both spatially incoherent representations and representations that jump from one position to another. We therefore developed state transition models to capture each of these dynamics.

@@

One of our criteria for a more optimal method is that it use all of the available spiking data. Using only clustered spikes discards all of the spike events that cannot be uniquely assigned to a putative single neuron, substantially reducing the quality of the resulting decoding. Additionally, spike sorting is not necessary to recover the underlying neural population dynamics \cite{TrautmannAccurateEstimationNeural2019}. We therefore adopted a "clusterless" approach which directly relates spikes and their multiunit spike waveform features to position without spike sorting, which we call clusterless decoding (see Methods). Clusterless decoding previously has been used to successfully identify theta sequences and replay sequences in the hippocampus (\cite{KloostermanBayesiandecodingusing2014, ChenTransductiveneuraldecoding2012,DengRapidclassificationhippocampal2016, KayConstantSubsecondCycling2020}). Applying a clusterless decoder to the same SWR, we get similar classification of the sequence (Figure \ref{2}B, D), both with 1D linearized position and 2D position (Figure 2-Supplemental 1B). The spatial extent of the event is longer and the estimate of the posterior probability of position is narrower for the clusterless model, however. This reflects the clusterless model's access to a larger pool of data that provides more information about the extent of the event, and more certainty in the latent position and the dynamic (Figure \ref{2}D vs C).

@@

Previous work focusing on spatially sequential replay reported that only a minority of events contain sequential spatial content [REFS]. We therefore asked what fraction of classified events contained spatially coherent content in any form, defined as a SWR containing any times with stationary, stationary-continuous mixture, or continuous dynamics (see Methods). We found that 22478 of 23382 or 96\% of classified SWRs had time points with spatially coherent structure. We then asked what fraction of events contained spatially incoherent content, defined as a SWR containing any times with fragmented-continuous mixture or fragmented. We found that only 3353 of 23382 or 14\% of classified SWRs had time points with spatially incoherent structure. This was consistent across animals (Figure \ref{3}F). We then asked what percentage of classified SWR contained continuous content, like would typically be analyzed when using the Bayesian decoder. Here our findings were consistent with previous reports: in our datasets 4511 of  23382 or 19\% of classified SWRs had some continuous content (Figure \ref{3}G). Thus, focusing on only continuous content misses a large fraction of events where there is evidence for spatially coherent, but not continuous content.

@@

Our model is specified in the context of a latent position associated with different dynamics. This specification allows us to not only decode events, but also to quantify the model's certainty in each position estimate at each moment in time given the model parameters. To do so we can compute the cumulative spatial extent of the 95\% highest posterior density (HPD) values of the posterior linear position. Larger values of the cumulative HPD indicate the model is less certain about position, because the HPD values are spread out over more of the track at a given time point, whereas lower values indicate that the model is more certain about the estimate of position, because the HPD values are more concentrated and cover less of the track. Thus, the HPD provides a complementary measures of spatial coherence, and evaluating it allows us to verify that the events we defined as spatially coherent correspond to events where there is high certainty around the position estimates.

We found that spatially coherent events did indeed have lower cumulative HPDs. Figure \ref{4}A and B show two example SWRs that were classified as having stationary and continuous dynamics, respectively. The bulk of the HPD values at each time step in these SWRs is concentrated in a small portion of the track and the cumulative HPD is low throughout the SWRs. In contrast, Figure \ref{4}C shows a SWR where the dynamics are fragmented and correspondingly, the HPD is much more spatially diffuse and the cumulative HPD is much higher. The HPD also provides insights into the moment-by-moment structure of each event, which can change over the time course of a SWR. An example of this change is shown in Figure \ref{4}D, where the cumulative HPD is small for most of the SWR until the end, where the uncertainty of position becomes much higher, reflecting a transition from a spatially coherent to a spatially incoherent representation. Overall, when we examined the average cumulative HPD for each SWR, grouped by dynamic, we found a clear bi-modal split between spatially coherent dynamics and spatially incoherent dynamics (Figure \ref{4}E). For the spatially coherent dynamics, the cumulative HPD was much lower than the spatially incoherent dynamics (median 24 cm, spatially coherent vs. median 238 cm, spatially incoherent).

@@

Our results indicate that most events with coherent spatial content are not continuous; that is, trajectories that move many times the running speed of the animal. We therefore examined these events in more detail. We first noted that most of the SWRs (16454 of 23382 classified SWRs or 70\%) were well described by a single dynamic. Surprisingly, the most common SWRs were categorized exclusively as stationary-continuous mixtures (9860 of 23382 classified SWRs or 42\%, Figure \ref{5}A). These events contain representations that move at rates more consistent with real-world movement (Fig. 1G), but were slightly shorter, on average, than events with continuous dynamics (median duration: stationary-continuous-mixture 72 ms vs. continuous 95 ms, Figure \ref{5}B). Nonetheless, both these slow-moving events and continuous events represented locations that were some distance away from the animal (mean trajectory distance from animal's position: stationary-continuous-mixture 51 cm vs. continuous 42 cm, Figure \ref{5}C). This indicates that the trajectories of these SWRs, like ones that have faster continuous dynamics, do not represent the animal's actual position.

The next most common category is those SWRs that were exclusively stationary (5149 of 23382 or 22\%). Unlike the stationary-continuous mixtures, most of these events represented a location close to the the animal's actual position (Figure \ref{5}C). There were, however, a number of stationary events  that represented positions far from the animal. We set a threshold of 30 cm distance from the animal to identify "non-local stationary trajectories" and found these trajectories in 1585 of 23382 (~7\%) of classified SWRs. About half of these (731 classified SWRs) were stationary throughout the duration of the SWR. These non-local stationary trajectories most commonly represented reward wells or choice points (Figure \ref{5}D), consistent with \cite{JaiDistincthippocampalcorticalmemory2017}, but a small portion of these trajectories occurred at other locations of the track. This suggests that these representations can be flexibly deployed to represent discrete locations.

Finally, we examined the overall activity levels across dynamics. We found that all the dynamics had similar average tetrode multiunit spiking rates (Figure \ref{5}E). This indicates that all dynamics, including the slower dynamics such as stationary and stationary-continuous mixtures, were driven by sustained spiking information and could not be explained by the absence of spiking.

\section*{Discussion}
Here we developed a state space model that identifies and quantifies generative activity in terms of a mixture of three movement dynamics: stationary, continuous, and fragmented. We showed that this model is interpretable; it has well-defined parameters and allows us to explain a large fraction of SWRs in terms of these dynamics---far greater than had been explained before. We showed that this model is flexible; it can be used with sorted or clusterless spikes, it can be used with 1D or 2D positions, and---because it uses a mixture of dynamics---it allowed us to confidently discover SWRs with non-constant speed and non-linear trajectories that would be difficult to characterize with the typical Bayesian decoder. Finally, we showed that this model is robust; it was relatively insensitive to transition parameters and depended strongly on the hippocampal place representation structure.

Using this model, we found that most SWRs contained spatially coherent content. Importantly, this content was not predominantly trajectories that occurred at many times the speed of the animal, but trajectories that were much closer to the animal's actual movement speed. Many of these events did not simply represent the animal's current location, but some other location on track.

Our results challenge the notion that hippocampal replays consist only of high speed events. Are stationary events during SWR that represent the animal's position different from stationary events that represent non-local positions

Our results complement those of Stella and colleagues \cite{StellaHippocampalReactivationRandom2019} who found that hippocampal replay trajectories followed Brownian like diffusion dynamics. We found that  We also clarify how these results can coexist with Pfeiffer and Foster \cite{PfeifferAutoassociativedynamicsgeneration2015} who found that a subset of replay trajectories alternated between stationary and jumpy representations. We find that, for our task, these jumpy representations were relatively rare.

Why our results are different from the Bayesian decoder
seek to describe as many SWR events, clusterless decoding, more permissive model with different dynamics, explicit descriptive model instead of implicit noisy model, smaller time bins

Our model is similar to the competing temporal models approach of Johnson and Redish 2008 \cite{JohnsonMeasuringdistributedproperties2008}, but allows for more . Our model also represents a middle ground between Hidden Markov-style models \cite{MaboudiUncoveringtemporalstructure2018, ChenBayesiannonparametricmethods2016, LindermanBayesiannonparametricapproach2016}, which seek to be environment-agnostic discovers of sequential patterns but require a lot of data, and typical Bayesian decoder, which has very restrictive assumptions about the nature of the trajectories. allows SWR to quickly be identified because of the discrete states. Can work in 2D. Can precisely identify timing of dynamics because of acausal decoding

We believe our model can enable a richer set of analyses that takes advantage of all the data, 
The model can easily be extend to incorporate other previously experienced environments by training place field models on those environments and including the appropriate movement transition matrices for those environments.

@@

\subsection*{Behavioral Task}
All animals performed a W-track spatial alternation task, which is described in detail in \cite{KarlssonAwakereplayremote2009}. In brief, each day, animals alternated between 20 minute rest epochs in a rest box and 15 minutes run epochs in the W-track. On the W-track, animals alternated between two types of runs: inbound and outbound. On inbound runs, the animal had to run from an outer arm to the center arm. On outbound runs, the animals had to run from the center arm to the least recently visited outer arm. Animals were rewarded for performing the task correctly. Only run epochs with at least 9 putative hippocampal pyramidal cells that fired at least 100 spikes were included in the analysis.

\subsection*{Position of the Animal and Linearization}
The animal's 2D position was estimated from digital video (30 Hz) of two infrared diodes placed on the head stage preamplifiers using a semi-automated analysis. In order to decrease the time it takes to run the model, the 2D position of the animal was converted into a 1D position. This is done by first defining a 2D graph representation of the track (herein referred to as the track graph), where edges correspond to segments of the W-track and nodes represent intersection points between those segments. Then, based on the algorithm in \cite{NewsonHiddenMarkovmap2009}, we use a Hidden Markov Model (HMM) to classify the most likely track segment the animal is on. Using the HMM takes into account the time dependent nature of the data and helps prevents sudden jumps from one track segment to another, which is particularly important near intersections. The observation model of the HMM is Gaussian and it models the likelihood of being on a track segment as the Gaussian distance to that segment with a 5 cm standard deviation. The state transition model is an empirically estimated state transition that changes with each time point that tries to ensure that the Euclidean distance between successive position estimates is similar to the shortest path distance along the graph between successive position estimates. A slight bias of 0.1 is given to the diagonal to encourage staying on the same track segment. The most likely track segment the animal is on is computed using the Viterbi algorithm. After finding the track segment that corresponds to each 2D position, the 2D position is projected onto the track segment. This allows us to define a distance from the center well in terms of shortest path length on the track, where 0 cm represents the center well position. The linear distance can then be converted into a linear position by assigning each track segment a position in 1D space. 15 cm gaps were placed between the center arm, left arm, and right arms in 1D space to prevent any smoothing done in the model from bleeding over inappropriately. The code used for linearization can be found at \url{https://github.com/Eden-Kramer-Lab/loren_frank_data_processing}.

\subsection*{Sorted Spikes, Multiunit Spikes and Waveform Features}
To obtain the neural data used for decoding, hippocampus electrical potentials were recorded at 30 kHz, referenced to a corpus callosum tetrode, and then digitally filtered between 600 Hz and 6 kHz. Events were considered threshold crossing events if any one of the four tetrode wires exceeded a 60 $\mu V$ threshold. These multiunit threshold crossing events and the associated peak amplitudes of the tetrodes wires---defined as the electrical potential value at the time of the maximum amplitude at the threshold crossing---were used as data for the clusterless version of the model.

@@

Lastly, we evaluate the likelihood of the observations $p(O_{k} \mid x_{k}, I_{k})$ based on an encoding model fit during the encoding period. We assume the likelihood is the same for each dynamic $I_k$, so we only need to evaluate $p(O_{k} \mid x_{k})$. It has been shown \cite{ZhangInterpretingNeuronalPopulation1998, BrownStatisticalParadigmNeural1998} that the Poisson likelihood with sorted spikes can be computed as:
$$
p(O_{k} \mid x_{k}) = p(\Delta N_{k}^{(1:C)} \mid x_{k}) \propto
\prod^{C}_{i=1} [\lambda_{i}(t_{k} \mid x_{k})\Delta_{k}]^{N_{t_{k}}^{i}} \\
    \exp[-\lambda_{i}(t_{k} \mid x_{k})\Delta_{k}]
$$
where $N_{t_{k}}^{i}$ represents a spike at time $t_k$ from cell $i$, $\Delta_k$ is the time bin size, and $\lambda_{i}(t_{k} \mid x_{k})$ is the instantaneous firing rate of cell $i$ given position $x_{k}$. $\lambda_{i}(t_{k} \mid x_{k})$ is the "place field" of the cell, which can be estimated by fitting a generalized linear model to each cell's spiking during the encoding period.

Likewise, it has been shown \cite{ChenTransductiveneuraldecoding2012, DengRapidclassificationhippocampal2016, KloostermanBayesiandecodingusing2014} that the clusterless likelihood can be computed as: 
$$
p(O_{k} \mid x_{k}) = p(\Delta N_{k}^{(1:E)}, \vec{m_k} \mid x_{k}) \propto
\prod^{E}_{i=1} [\lambda_{i}(t_{k}, \vec{m}_k \mid x_{k})\Delta_{k}]^{N_{t_{k}}^{i}} \\
    \exp[-\Lambda_{i}(t_{k} \mid x_{k})\Delta_{k}]
$$
where $\lambda_{i}(t_{k}, \vec{m}_k \mid x_{k})$ is now a generalized firing rate that depends on an associated wave form features $\Vec{m}$ and $\Lambda_{i}(t_{k} \mid x_{k})$ is a marginal firing rate that is equivalent to a place field estimated on multiunit spikes. Both of these rates can be defined as:
$$
\lambda_{i}(t_{k}, \vec{m}_k \mid x_{k}) = \mu_i \frac{p_{i}(x_k, \vec{m}_k)}{\pi(x)}
$$
and
$$
\Lambda_{i}(t_{k} \mid x_{k}) = \mu_i \frac{p_{i}(x_k)}{\pi(x)}
$$
where $\mu_i$ is the mean firing rate for tetrode $i$, $\pi(x)$ is the spatial occupancy of the animal on the track, and $p_{i}(x_k, \vec{m}_k)$ and $p_{i}(x_k)$ are the joint probability of observing waveform features $\Vec{m}$ and position $x$ and the probability of observing position $x$, respectively. $\pi(x)$, $p_{i}(x_k, \vec{m}_k)$, $p_{i}(x_k)$ can all be estimated by training a kernel density estimator on each tetrode's spike waveform features and corresponding position during the encoding period.

\subsection*{Encoding - Sorted Spikes}
In order to encode how each cell's spiking activity relates to position (the place field), we fit a generalized linear model (GLM) with a Poisson response distribution to each cell's spiking activity during the encoding period, which we define as all movement times (time periods when the running speed is greater than 4 cm/s). We estimate the parameters $\beta$, which consist of $\beta_{0}$, the average firing rate over time, and $\beta_{i}$, weights for third degree B-splines basis functions $f_{i}$ over position (or tensor products of the B-splines when position is two dimensional). B-spline basis functions are used because place field firing activity is assumed to vary smoothly over position and this prior knowledge can be exploited to reduce the total number of model parameters needed. Each basis function is spaced every 5 cm over the range of the position and zero constrained so that the change encoded by the parameters is relative to the baseline firing rate. We use a log link function to convert the linear combination of parameters to an instantaneous firing rate over time $\lambda(t)$ to ensure the rate is always positive.

@@

A small L2 penalization term $-\lambda\|\beta_{i}\|_{2}^{2}$ used to prevent model fitting instability when spiking activity is very low. We set this to 0.5 for all cells. Fitting is done by maximizing the penalized likelihood using a Newton-Raphson algorithm. The code used to fit the GLMs is available at \url{https://github.com/Eden-Kramer-Lab/regularized_glm}.

\subsection*{Encoding - Clusterless}
In order to encode how each tetrode's unsorted spiking activity and waveform features relate to position, we used kernel density estimation (KDE) to estimate the following distributions: $\pi(x)$, $p_{i}(x_k)$, and $p_{i}(x_k, \vec{m}_k)$. We used KDEs of the form: 
$$
kde(y) = \frac{1}{N h_1...h_D} \sum^{N}_{i=1} \prod^{D}_{d=1} K_{d}\left(\frac{y_d - y_{i,d}}{h_d}\right)
$$
where $y$ is the data with $D$ dimensions, $K$ is a 1 dimensional Gaussian kernel with bandwidth $h_d$, $y_{i, d}$ is the data observed during movement, $N$ is the number of observations during the movement period. For $\pi(x)$, $y_{i, d}$ is all positions observed during movement. For $p_{i}(x_k)$, $y_{i, d}$ is all positions at the time of multiunit spikes during movement. For $p_{i}(x_k, \vec{m}_k)$, $y_{i, d}$ is all positions at the time of multiunit spikes and their associated waveform features during movement. We choose the bandwidth $h_d$ to be 6.0 cm for any position dimension and 24.0 $\mu V$ for any spike amplitude dimension, because these parameters minimized decoding error in \cite{KloostermanBayesiandecodingusing2014}. The mean firing rate $\mu_k$ was also estimated during movement.

\subsection*{Decoding}
In order to decode $p(x_{k}, I_{k} \mid O_{1:T})$, we used a grid based approximation of the latent position $x_k$ that respected the geometry of the track. For 1D linearized positions, we discretized the position space based on the same track graph we used for linearization by finding bins less than or equal to the chosen position bin size for each edge of the graph. For 2D positions, we discretized the position space by binning 2D positions occupied by the animal with equal sized bins of the chosen position bin size, followed by morphological opening to get rid of any holes smaller than the bin size. This was done using Scipy image processing tools \cite{SciPy1.0ContributorsSciPyfundamentalalgorithms2020}. This grid based approximation allows us to use Reimann sums to approximate the integrals in the causal filter and acausal smoother equations. We chose a position grid with bins of 3 cm in width (and height if the model was computed with 2D positions) in order to get good resolution for the random walk transition matrices (which had 6 cm variance) as well as for the clusterless and sorted spikes decoding (which have 6 cm bandwidth for the KDE position dimensions and 5 cm spline knots for the GLM). For sorted spikes decoding, we evaluated the place field $\lambda_{i}(t_{k} \mid x_{k})$ on the midpoint of these bins. Likewise, for clusterless decoding, we evaluated the spike amplitudes observed during the decoding period by evaluating the KDE for $p_{i}(x_k, \vec{m}_k)$ for the midpoint of these bins.

We also used this grid bins in combination with the track graph in order to construct the appropriate 1D random walk transition matrices that respected the track geometry. To do this, we inserted the bin centers as nodes in the track graph, and then computed the shortest path distance between all pairs of position bin centers. We then evaluated a zero mean Gaussian with a variance of 6 cm on these distances to get the appropriate transition probability of moving from bin to bin.
